---
title: "GIS in R - exercises"
author: "R Félix"
date: "MQAT 2023"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    # code_folding: "hide"
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(sf)
library(mapview)
```


# Intro

This Rmarkdown notebook aims to show some examples of how to solve the classroom proposed exercises in QGIS, but in R.

There are several ways to reach the same solution. Here we present only one of them.

# Represent Transport Zones

Download and open `TRIPSgeo_mun.gpkg` and `TRIPSgeo_freg.gpkg` under [MQAT/geo/](https://github.com/U-Shift/MQAT/tree/main/geo) repository.

```{r getdata1}
TRIPSgeo_mun = st_read("geo/TRIPSgeo_mun.gpkg", quiet = TRUE) # we add quiet = TRUE so we don't get annoying messages on the info
TRIPSgeo_freg = st_read("geo/TRIPSgeo_freg.gpkg", quiet = TRUE)

# you can also open directly from url from github. example:
# TRIPSgeo_mun = st_read("https://github.com/U-Shift/MQAT/raw/main/geo/TRIPSgeo_mun.gpkg")
```

Represent Transport Zones with Total, and with Car %.

```{r carper}
# create Car_per variable
TRIPSgeo_mun = TRIPSgeo_mun |> mutate(Car_per = Car / Total * 100)
TRIPSgeo_freg = TRIPSgeo_freg |> mutate(Car_per = Car / Total * 100)

#Vizualize in map
mapview(TRIPSgeo_mun, zcol = "Car_per")
mapview(TRIPSgeo_freg, zcol = "Car_per", col.regions = rev(hcl.colors(9, "-Inferno"))) #palete inferno com 9 classes, reverse color ramp
```

# Centroids

## Geometric centroids

```{r geocentroid}
CENTROIDSgeo = st_centroid(TRIPSgeo_mun)
# mapview(CENTROIDSgeo)
```


## Weigthed centroids

Get BGRI Data^[Base Geográfica de Referenciação de Informação, Censos 2021] from INE website, at *Área Metropolitana de Lisboa* level: [https://mapas.ine.pt/download/index2021.phtml](mapas.ine.pt/download/index2021.phtml)

```{r getcensus}
BGRI = st_read("original/BGRI21_LISBOA.gpkg", quiet = TRUE)
```

It is not that easy to estimate weighted centroids with R. See [here](https://wzbsocialsciencecenter.github.io/spatially_weighted_avg/).  
We will make a bridge connection to QGIS to use its native function of mean coordinates.  


```{r weighted}
library(qgisprocess)
# qgis_search_algorithms("mean") # search the exact function name
# qgis_get_argument_specs("native:meancoordinates") |> select(name, description) # see the required inputs

# with population
CENTROIDSpop = qgis_run_algorithm(algorithm = "native:meancoordinates",
                                  INPUT = BGRI,
                                  WEIGHT = "N_INDIVIDUOS",
                                  UID = "DTMN21")
CENTROIDSpop = st_as_sf(CENTROIDSpop)

# with buildings
CENTROIDSbuild = qgis_run_algorithm(algorithm = "native:meancoordinates",
                                  INPUT = BGRI,
                                  WEIGHT = "N_EDIFICIOS_CLASSICOS",
                                  UID = "DTMN21")
CENTROIDSbuild = st_as_sf(CENTROIDSbuild)
```


## Compare in map

```{r mapcentroid}
mapview(CENTROIDSgeo) + mapview(CENTROIDSpop, col.region = "red") + mapview(CENTROIDSbuild, col.region = "black")
```

See how the building, poulation and geometric centroids of Montijo are appart, from closer to Tagus, to the rural area.


# Desire Lines

Download `TRIPSdl_mun.gpkg` 

```{r getdl}
TRIPSdl_mun = st_read("geo/TRIPSdl_mun.gpkg", quiet = TRUE) 
```

Filter intrazonal trips, and trips with origin or desination in Lisbon.

```{r withlx}
TRIPSdl_mun = TRIPSdl_mun |> 
  filter(Origin_mun != Destination_mun) |> 
  filter(Total > 5000) # remove noise

mapview(TRIPSdl_mun, zcol = "Total", lwd = 5)
```

```{r filterlx}
TRIPSdl_mun_noLX = TRIPSdl_mun |> 
  filter(Origin_mun != "Lisboa", Destination_mun != "Lisboa")

mapview(TRIPSdl_mun_noLX, zcol = "Total", lwd = 8)
```

You can replace the `Total` with other variable, such as `Car`, `PTransit`, and so on.

> Note: The function [`od_oneway()`](https://docs.ropensci.org/stplanr/reference/od_oneway.html) aggregates oneway lines to produce bidirectional flows. By default, it returns the sum of each numeric column for each bidirectional origin-destination pair. This is better for viz purpouses.

# Euclidean vs. Routing distance

## Euclidean distance

### Create new point at IST

```{r createist}
IST = st_sfc(st_point(c(-9.1397404, 38.7370168)), crs = 4326)
```

### Import survey and visualize

```{r survey}
SURVEY = read.csv("geo/SURVEYist.txt", sep = "\t") # tab delimiter
SURVEY = st_as_sf(SURVEY, coords = c("lon", "lat"), crs = 4326) # transform as geo data

mapview(SURVEY, zcol = "MODE") + mapview(IST, col.region = "red", cex = 12)
```

### Reproject layers

In R we can process distances in meters on-fly.

Buy here is the code to project layers from Geographic coordinates (WGS 84 - EPSG:[4364](https://epsg.io/4326)) to Projected coordinates (Pseudo-Mercator - EPSG:[3857](https://epsg.io/3857), or Portuguese Tranversor-Mercator 06 - EPSG:[3763](https://epsg.io/3763)).

```{r projectlayers}
ISTprojected = st_transform(IST, crs = 3857)
SURVEYprojected = st_transform(SURVEY, crs = 3857)
```

### Straight lines and distance

Nearest point between the two layers. As we only have 1 point at IST layer, we will have the same number of lines as number of surveys = `r nrow(SURVEY)`.

```{r eucdistance}
SURVEYeuclidean = st_nearest_points(SURVEY, IST, pairwise = TRUE) |> st_as_sf() # this creates lines

mapview(SURVEYeuclidean)

SURVEY$distance = st_length(SURVEYeuclidean) # compute distance and add directly in the first layer

summary(SURVEY$distance) # in meters

# to remove the units - can be useful
SURVEY$distance = units::drop_units(SURVEY$distance)
```

The same function can be used to find the closest GIRA station to each survey home location. And also check where are the ones that are far away from GIRA.

```{r girahub}
GIRA = st_read("geo/GIRA2023.geojson", quiet = TRUE) # we can also read geojson with this function!

nearest = st_nearest_feature(SURVEY, GIRA) # creates an index of the closest GIRA station id

SURVEY$distanceGIRA = st_distance(SURVEY, GIRA[nearest,], by_element = TRUE)

mapview(SURVEY, zcol = "distanceGIRA") +
  mapview(GIRA, col.regions = "grey20", cex = 4, legend = FALSE)
```

## Routing distance

We use the [openrouteservice-r](https://giscience.github.io/openrouteservice-r/) package. For that you need to [create an account](https://openrouteservice.org/dev/#/signup) and get a Token / api key.

> **Note on how to store credentials in R**  
Using an API key from [OpenRouteService](https://openrouteservice.org/dev/#/signup), you should store it at your computer and **never show it directly on code**.  
For that `usethis::edit_r_environ()` and paste your token as `ORS_API_KEY="xxxxxxxxxxxxxxxxxxxxxx"` (replace with your token).    
Save the .Renviron file, and press `Ctrl+Shift+F10` to restart R so it can take effect.

See the [documentation](https://giscience.github.io/openrouteservice-r/articles/openrouteservice.html) for more details.

### Distances 1 point to many points

Estimate the time and distance by `foot-waking` and `driving-car`, `fastest` mode, from survey locations (under 2 km^[for speed-up purposes - api request limit up to 40 / minute]) to IST.

```{r routingdataprep}
# devtools::install_github("GIScience/openrouteservice-r")
library(openrouteservice)
# ors_api_key(Sys.getenv("ORS_API_KEY")) # one time setup

# get coordinates variable
SURVEY$coordinates = st_coordinates(SURVEY)
IST$coordinates = st_coordinates(IST)

# Filter only the locations up to 2km euclidean
SURVEYsample = SURVEY |> filter(distance <= 2000)
# nrow(SURVEYsample) # 95
```


Although it is the same algorithm, here it works differently from QGIS.

There are many ways of doing this. If we want to know only time and distance, and **not the route** itself, we can use the `ors_matrix()`. See example [here](https://web.tecnico.ulisboa.pt/~rosamfelix/gis/trips/timedistancematrix.html#Distance_and_time_matrix).  
If we need the route, we should use the function `ors_directions()`. This one is not that easy to set-up because the function is prepared to retrieve only one result per request :( So we do a loop. Don't worry, it is not that

```{r orsloop1, eval=FALSE, include=TRUE}
ROUTES_foot = data.frame() # initial empty data frame

# loop - the origin (i) is the survey location, and the IST is always the same destination
for (i in 1:nrow(SURVEYsample)) {
  ROUTES1 = ors_directions(
    data.frame(
      lon = c(SURVEYsample$coordinates[i, 1], IST$coordinates[1, 1]),
      lat = c(SURVEYsample$coordinates[i, 2], IST$coordinates[1, 2])
    ),
    profile = "foot-walking", # or driving-car cycling-regular cycling-electric
    preference = "fastest", # or shortest
    output = "sf"
  )
  ROUTES1$distance = ROUTES1$summary[[1]]$distance # extract these values from summary
  ROUTES1$duration = ROUTES1$summary[[1]]$duration
  
  ROUTES_foot = rbind(ROUTES_foot, ROUTES1) # to keep adding in the same df
}

ROUTES_foot = ROUTES_foot |>
  select(distance, duration, geometry) |> # discard unnecessary variables
  mutate(ID = SURVEYsample$ID) # cbind with syrvey ID
```

Repeat the same for `car-driving`.

```{r orsloop2, eval=FALSE, include=TRUE}
ROUTES_car = data.frame() # initial empty data frame

# loop - the origin (i) is the survey location, and the IST is always the same destination
for (i in 1:nrow(SURVEYsample)) {
  ROUTES1 = ors_directions(
    data.frame(
      lon = c(SURVEYsample$coordinates[i, 1], IST$coordinates[1, 1]),
      lat = c(SURVEYsample$coordinates[i, 2], IST$coordinates[1, 2])
    ),
    profile = "driving-car", # or cycling-regular cycling-electric foot-walking
    preference = "fastest", # or shortest
    output = "sf"
  )
  ROUTES1$distance = ROUTES1$summary[[1]]$distance # extract these values from summary
  ROUTES1$duration = ROUTES1$summary[[1]]$duration
  
  ROUTES_car = rbind(ROUTES_car, ROUTES1) # to keep adding in the same df
}

ROUTES_car = ROUTES_car |>
  select(distance, duration, geometry) |> # discard unnecessary variables
  mutate(ID = SURVEYsample$ID) # cbind with syrvey ID
```

```{r importexport1, include=FALSE}
# st_write(ROUTES_foot, "original/routes_foot.geojson")
# st_write(ROUTES_car, "original/routes_car.geojson")

ROUTES_foot = st_read("original/routes_foot.geojson", quiet = TRUE)
ROUTES_car = st_read("original/routes_car.geojson", quiet = TRUE)
```

## Compare distances

We can compare the euclidean and routing distances that we estimated for the survey locations under 2 km.

```{r distancessummary}
summary(SURVEYsample$distance) # Euclidean
summary(ROUTES_foot$distance) # Walk
summary(ROUTES_car$distance) # Car
```

## Vizualise routes

Visualize with transparency of 30%

```{r maproutes}
mapview(ROUTES_foot, alpha = 0.3)
mapview(ROUTES_car, alpha = 0.3, color = "red")
```

We can also use the `overline()` [function from stplanr package](https://docs.ropensci.org/stplanr/reference/overline.html) to break up the routes when they overline, and add them up.

```{r overline, message=FALSE, warning=FALSE}
library(stplanr)

# we create a value that we can later sum, it also could be the number of trips represented by this route. in this case is only one respondent per route
ROUTES_foot$trips = 1 

ROUTES_foot_overline = overline(
  ROUTES_foot,
  attrib = "trips",
  fun = sum
)

mapview(ROUTES_foot_overline, zcol = "trips", lwd = 3)
```

*   How many people are entering IST by the stairs near *Bar de Civil*?  
*   And by the North gate?
*   And from Alameda stairs?

# Buffers vs. Isochones and Service Areas

## Buffer

Represent a buffer of 500 m and 2000 m from IST^[Here I selected only the first variable because now we also have the coordinates information (unecessary for this procedure)].

```{r bufferIST}

# BUFFERist500 = st_buffer(IST, dist = 500) # non  projected - results may be weird
BUFFERist500 = geo_buffer(IST[1], dist = 500) # from stplnar, to make sure it is in meters.
BUFFERist2000 = geo_buffer(IST[1], dist = 2000)

mapview(BUFFERist500) + mapview(BUFFERist2000, alpha.regions = 0.5)
```

## Isochrone

### Isochrone from 1 point - distance

We use again the `openrouteservice` r package.

```{r isoch1}
ISOCist = ors_isochrones(
  IST$coordinates,
  profile = "foot-walking",
  range_type = "distance", # or time
  range = c(500, 1000, 2000),
  output = "sf"
)

ISOCist = arrange(ISOCist, -value) # to make the larger polygons on top of the table so the are displayed behind.

mapview(ISOCist, zcol = "value", alpha.regions = 0.5)
```

As you can see, the distance buffer of 500m is larger than the isochrone of 500m.
Actually we can measure their area of reach.

```{r arearatio}
ISOCist$area = st_area(ISOCist)
BUFFERist500$area = st_area(BUFFERist500)
BUFFERist2000$area = st_area(BUFFERist2000)

ratio1 = BUFFERist500$area / ISOCist$area[ISOCist$value == 500] # 1.71
ratio2 = BUFFERist2000$area / ISOCist$area[ISOCist$value == 2000] # 1.22
```

The euclidean buffer of 500m is `r round(ratio1, 2)` times larger than its isochrone, and the buffer of 2000m is `r round(ratio2, 2)` times larger than its isochrone.


### Isochrone from more than 1 point - time

For this purpose we will use the high schools dataset.

```{r getschools}
# import schools
SCHOOLS = st_read("geo/SCHOOLS_basicsec.gpkg", quiet = TRUE)

SCHOOLS$coordinates = st_coordinates(SCHOOLS) # create coordinate variable

SCHOOLShigh = SCHOOLS |> filter(Nivel == "Secundario") # filter the high schools

```

And proceed with the time isochrones, for a range of 20 min, with 5 min intervals.

```{r isoch2, eval=FALSE, include=TRUE}
coor = data.frame(lon = SCHOOLShigh$coordinates[,1], lat = SCHOOLShigh$coordinates[,2])
coor_max5 = sample_n(coor, 5) # error of api, get a loop to overpass this!

ISOCist_5 = ors_isochrones(
  coor_max5,
  profile = "foot-walking",
  range_type = "time", # or distance
  range = 20*60, # 20 minutes in seconds
  interval = 5*60, # to have intervals of 5 minutes
  output = "sf"
)
```

Because `openrouteservece` only allows a [max of 5 requests](https://openrouteservice.org/restrictions/) for isochrones at a time, we put it in a loop to run for all the 20 high schools.

```{r isoch2loop}
for (i in 1:nrow(coor)) {
  ISOCist_i =
  tryCatch(
  ors_isochrones(
    coor[i, ],
    profile = "foot-walking",
    range_type = "time", # or distance
    range = 20*60, # 20 minutes in seconds
    interval = 5*60, # to have intervals of 5 minutes
    output = "sf"
  ), 
  error=function(e){cat("Error: Openrouteservice API request failed")} # trick to ignore the ones that return errors
  )
  ISOCist = rbind(ISOCist, ISOCist_i)
}

ISOCist = arrange(ISOCist, -value) # to make the larger polygons on top of the table so the are displayed behind.

ISOCist = arrange(ISOCist, -value) # to make the larger polygons on top of the table so the are displayed behind.

mapview(ISOCist, zcol = "value", alpha.regions = 0.5)
```

```{r}

```

*Work In Progress*