---
title: "Trips in AML with maps"
author: "R Félix"
date: "MQAT 2023"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: "hide"
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(stplanr)
library(tmap)
tmap_mode("view")
```


## Get OD data from IMOB

The [**biclaR**](https://biclar.tmlmobilidade.pt/) project hosts several [open datasets](https://github.com/U-Shift/biclar/releases/tag/0.0.1) useful for this spatial analysis relateg with transportation in the Lisbon Metro region.

```{r readdata}
TRIPSall = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))

TRIPSall[TRIPSall == "Setubal"] = "Setúbal" #rename

knitr::kable(head(TRIPSall)) # have a look at data
```

Remove **intrazonal** trips and group by Private vs Public

```{r tripsclean}
# With all modes
TRIPSinter = TRIPSall |>
  filter(Origem != Destino) |> # remove internal trips
  as.data.frame() |> 
  group_by(Origem, Destino) |> 
  summarise(viagens = sum(viagens)) |> 
  ungroup() |> 
  od_oneway() # sum the O with D - doesn't matter the start and end point for vizualization purpouses

  
# With variables of public or private modes
TRIPSvs = TRIPSall |>
  filter(Origem != Destino) |> # remove internal trips
  as.data.frame() |> 
  mutate(Tipo = recode(modo,  # recode in TI and TC
         Bike = "TIndividual",
         Car = "TIndividual",
         Motorcycle = "TIndividual",
         Other = "TIndividual",
         Transit = "TColetivo",
         Walk = "TIndividual")) |> 
  group_by(Origem, Destino, Tipo) |> 
  summarise(viagens = sum(viagens)) |> 
  ungroup() |> 
  pivot_wider(names_from = Tipo,
              values_from = viagens)

TRIPSvs[is.na(TRIPSvs)] = 0 # há casos em que não há viagens registadas entre municípios
TRIPSvs = od_oneway(TRIPSvs) # magia para somar O-D com D-O


knitr::kable(head(TRIPSvs)) # viagens separadas por TI e TC
```

A total of `r round(sum(TRIPSall$viagens)/1000000,1)` million trips are made daily in the AML, of which `r round(sum(TRIPSinter$viagens)/1000)` thousand are intercity.

## Get AML maps

```{r amlgeo}
MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
MUNICIPIOSgeo$Concelho[MUNICIPIOSgeo$Concelho == "Setubal"] = "Setúbal"

AML_dtcc = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/ZONAMENTO_imob.Rds"))
AML_dtcc = AML_dtcc |> select(DTMN, DTMN_DSG) |> unique()

CENTROIDS = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/CENTROIDS_municipios.Rds"))
CENTROIDS = CENTROIDS |>
  left_join(AML_dtcc, by = c("DTMN11" = "DTMN")) |>
  select(DTMN_DSG, geometry) |>
  filter(DTMN_DSG != "Montijo")

MONTIJO = sf::st_read("geo/Montijo.geojson", quiet = TRUE) # Porque o centro das duas parts do Montijo não calha dentro do Montijo
CENTROIDS = rbind(CENTROIDS, MONTIJO)
```

## Make maps

Use `stplanr` to create the [desire lines](https://docs.ropensci.org/stplanr/reference/od2line.html).

```{r od2line}
LINHASod = od2line(flow = TRIPSinter,
                   zones = CENTROIDS,
                   zone_code = "DTMN_DSG") |> 
  filter(viagens >= 1000)
LINHASod_noLX = LINHASod |> filter(Origem != "Lisboa" & Destino != "Lisboa")

LINHASod_TI = od2line(flow = TRIPSvs |> select(-TColetivo),
                      zones = CENTROIDS,
                      zone_code = "DTMN_DSG") |> 
  filter(TIndividual > 1000)
LINHASod_TI_noLX = LINHASod_TI |> filter(Origem != "Lisboa" & Destino != "Lisboa")

LINHASod_TC = od2line(flow = TRIPSvs |> select(-TIndividual),
                      zones = CENTROIDS,
                      zone_code = "DTMN_DSG") |> 
  filter(TColetivo > 1000)
LINHASod_TC_noLX = LINHASod_TC |> filter(Origem != "Lisboa" & Destino != "Lisboa")
```

### All Trips

#### Including Lisbon

```{r intermunicipaisALL, out.width = '100%'}
m1 =  tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod %>% filter(viagens >=5000)) +
  tm_lines("viagens",
           palette = viridis::magma(n = 4, direction = -1),
           breaks = c(5000, 10000, 50000, 100000, 161000),
           lwd = "viagens",
           scale = 30,
           title.col = "Viagens intermunicipais") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m1
```

#### Excluding Lisbon

```{r intermunicipaissemLX, out.width = '100%'}
m2 = tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod_noLX %>% filter(viagens >=3000)) +
  tm_lines("viagens",
           palette = viridis::magma(n = 4, direction = -1),
           breaks = c(3000, 10000, 30000, 60000, 73000),
           lwd = "viagens",
           scale = 30,
           title.col = "Viagens intermunicipais, fora de Lisboa") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m2
```

### Only Public Transit trips

#### Including Lisbon

```{r intermunicipais_TC, out.width = '100%'}
m3 = tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod_TC %>% filter(TColetivo >=3000)) +
  tm_lines("TColetivo",
           palette = viridis::viridis(n = 4, direction = -1),
           breaks = c(3000, 5000, 10000, 25000, 50000),
           lwd = "TColetivo",
           scale = 30,
           title.col = "Viagens intermunicipais, em Transporte Coletivo") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m3
```

#### Excluding Lisbon

```{r intermunicipaissemLX_TC, out.width = '100%'}
m4 = tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod_TC_noLX %>% filter(TColetivo >=2000)) +
  tm_lines("TColetivo",
           palette = viridis::viridis(n = 4, direction = -1),
           breaks = c(2000, 4000, 7000, 10000, 13000),
           lwd = "TColetivo",
           scale = 30,
           title.col = "Viagens intermunicipais, em Transporte Coletivo (fora Lx)") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m4
```


### Only Private trips

#### Including Lisbon

```{r intermunicipais_TI, out.width = '100%'}
m5 = tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod_TI %>% filter(TIndividual >=5000)) +
  tm_lines("TIndividual",
           palette = viridis::plasma(n = 4, direction = -1),
           breaks = c(5000, 20000, 50000, 100000, 115000),
           lwd = "TIndividual",
           scale = 30,
           title.col = "Viagens intermunicipais, em Transporte Individual") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m5
```

#### Excluding Lisbon

```{r intermunicipaissemLX_TI, out.width = '100%'}
m6 = tm_shape(MUNICIPIOSgeo) +
  tm_borders(col = "grey") +
  tm_shape(LINHASod_TI_noLX %>% filter(TIndividual >=3000)) +
  tm_lines("TIndividual",
           palette = viridis::plasma(n = 4, direction = -1),
           breaks = c(3000, 10000, 25000, 50000, 65000),
           lwd = "TIndividual",
           scale = 30,
           title.col = "Viagens intermunicipais, em Transporte Individual (fora Lx)") +
  tm_shape(CENTROIDS) +
  tm_text("DTMN_DSG", size = 0.8, col = "gray25")
m6
```
