---
number-depth: 3
code-fold: false
format:
  pdf:
    prefer-html: true
---

# Euclidean and routing distances

We will show how to estimate euclidean distances (*as crown flights*) using `sf` package, and the distances using a road network using `openrouteservice` package.

## Euclidean distances

Taking the survey respondents' location, we will estimate the distance to the university (IST) using the `sf` package.

### Import survey data frame convert to sf

We will use a survey dataset with 200 observations, with the following variables: ID, Affiliation, Age, Sex, Transport Mode to IST, and latitude and longitude coordinates.

```{r}
#| message: false
library(dplyr)

SURVEY = read.csv("../data/SURVEY.txt", sep = "\t") # tab delimiter
names(SURVEY)
```

As we have the coordinates, we can convert this data frame to a spatial feature, as explained in the [Introduction to spatial data](spatial-data.qmd#create-spatial-data-from-coordinates) section.

```{r}
#| message: false
library(sf)

SURVEYgeo = st_as_sf(SURVEY, coords = c("lon", "lat"), crs = 4326) # convert to as sf data
```

### Create new point at the university

Using coordinates from Instituto Superior Técnico, we can directly create a simple feature and assign its crs.

```{r}
UNIVERSITY = data.frame(place = "IST",
                        lon = -9.1397404,
                        lat = 38.7370168) |>  # first a dataframe
  st_as_sf(coords = c("lon", "lat"), # then a spacial feature
           crs = 4326)
```

Visualize in a map:

```{r}
#| fig-format: png
#| message: false
library(mapview)
mapview(SURVEYgeo, zcol = "MODE") + mapview(UNIVERSITY, col.region = "red", cex = 12)
```

### Straight lines

First we will create lines connecting the survey locations to the university, using the `st_nearest_points()` function.

This function finds returns the nearest points between two geometries, and creates a line between them.
This can be useful to find the nearest train station to each point, for instance.

As we only have 1 point at UNIVERSITY layer, we will have the same number of lines as number of surveys = `r nrow(SURVEY)`.

```{r}
#| warning: false
#| fig-format: png
SURVEYeuclidean = st_nearest_points(SURVEYgeo, UNIVERSITY, pairwise = TRUE) |>
  st_as_sf() # this creates lines

mapview(SURVEYeuclidean)
```

Note that if we have more than one point in the second layer, the `pairwise = TRUE` will create a line for each combination of points.
Set to `FALSE` if, for instance, you have the same number of points in both layers and want to create a line between the corresponding points.

### Distance

Now we can estimate the distance using the `st_length()` function.

```{r}
#| eval: false
# compute the line length and add directly in the first survey layer
SURVEYgeo = SURVEYgeo |> 
  mutate(distance = st_length(SURVEYeuclidean))

# remove the units - can be useful
SURVEYgeo$distance = units::drop_units(SURVEYgeo$distance) 
```

We could also estimate the distance using the `st_distance()` function **directly**, although we would not get and sf with lines.

```{r}
SURVEYgeo = SURVEYgeo |> 
  mutate(distance = st_distance(SURVEYgeo, UNIVERSITY)[,1] |>  # in meters
           units::drop_units()) |>  # remove units
  mutate(distance = round(distance)) # round to integer

summary(SURVEYgeo$distance)
```

`SURVEYgeo` is still a points' sf.

## Routing Engines

There are different types of routing engines, regarding the type of network they use, the type of transportation they consider, and the type of data they need.
We can have:

-   Uni-modal vs. Multi-modal

    -   One mode per trip vs. One trip with multiple legs that can be made with different modes
    -   Multi-modal routing may require GTFS data (realistic Public Transit)

-   Output level of the results

    -   Routes (1 journey = 1 route)
    -   Legs
    -   Segments

-   Routing profiles

    -   Type of user
    -   fastest / shortest path
    -   avoid barriers / tolls, etc

![Routing options in [OpenRouteService](https://maps.openrouteservice.org/)](images/clipboard-1370155720.png){fig-align="center"}

-   Local vs. Remote (service request - usually web API)

    -   Speed vs. Quota limits / price
    -   Hard vs. Easy set up
    -   Hardware limitations in local routing
    -   Global coverage in remote routing, with frequent updates

Examples: [OSRM](https://project-osrm.org/), [Dodgr](https://urbananalyst.github.io/dodgr/), [r5r](https://ipeagit.github.io/r5r/), [Googleway](https://symbolixau.github.io/googleway/reference/access_result.html), [CycleStreets](https://m.cyclestreets.net/journey), [HERE](https://munterfi.github.io/hereR/).

<!-- ### Routing distances with `r5r` -->

<!-- We use the `r5r` package to estimate the distance using a road network [@Pereira2021r5r]. -->

<!-- ::: {.callout-note appearance="simple"} -->

<!-- To properly the setup r5r model for the area you are working on, you need to download the **road network** data from OpenStreetMap and, if needed, add a **GTFS** and **DEM** file, as it will be explained in the [next section](r5r.qmd). -->

<!-- ::: -->

<!-- ```{r} -->

<!-- #| message: false -->

<!-- #| warning: false -->

<!-- #| include: false -->

<!-- options(java.parameters = "-Xmx4G") # allocate memory -->

<!-- url = "https://github.com/U-Shift/EITcourse/releases/download/2024.09/network.dat" -->

<!-- download.file(url, destfile = "../data/network.dat", mode = "wb") -->

<!-- library(r5r) -->

<!-- r5r_network = setup_r5("../data/", overwrite = FALSE) # load existing network -->

<!-- ``` -->

<!-- We will use only respondents with a distance to the university less than 2 km. -->

<!-- ```{r} -->

<!-- #| warning: false -->

<!-- SURVEYsample = SURVEYgeo |> filter(distance <= 2000) -->

<!-- nrow(SURVEYsample) -->

<!-- ``` -->

<!-- We need an id (unique identifier) for each survey location, to be used in the routing functions of `r5r`. -->

<!-- ```{r} -->

<!-- # create id columns for both datasets -->

<!-- SURVEYsample = SURVEYsample |>  -->

<!--   mutate(id = c(1:nrow(SURVEYsample))) # from 1 to the number of rows -->

<!-- UNIVERSITY = UNIVERSITY |>  -->

<!--   mutate(id = 1) # only one row -->

<!-- ``` -->

<!-- #### Distances by car -->

<!-- Estimate the routes with time and distance by car, from survey locations to University. -->

<!-- ```{r} -->

<!-- #| warning: false -->

<!-- SURVEYcar = detailed_itineraries( -->

<!--   r5r_core = r5r_network, -->

<!--   origins = SURVEYsample, -->

<!--   destinations = UNIVERSITY, -->

<!--   mode = "CAR", -->

<!--   all_to_all = TRUE # if false, only 1-1 would be calculated -->

<!-- ) -->

<!-- names(SURVEYcar) -->

<!-- ``` -->

<!-- The [`detailed_itineraries()`](https://ipeagit.github.io/r5r/reference/detailed_itineraries.html) function is super detailed! -->

<!-- ::: {.callout-note appearance="simple"} -->

<!-- If we want to know only time and distance, and **not the route** itself, we can use the [`travel_time_matrix()`](https://ipeagit.github.io/r5r/reference/travel_time_matrix.html). -->

<!-- ::: -->

<!-- #### Distances by foot -->

<!-- Repeat the same for `WALK`[^4]. -->

<!-- [^4]: For bike you would use `BICYCLE`. -->

<!-- ```{r} -->

<!-- #| warning: false -->

<!-- #| code-fold: true -->

<!-- SURVEYwalk = detailed_itineraries( -->

<!--   r5r_core = r5r_network, -->

<!--   origins = SURVEYsample, -->

<!--   destinations = UNIVERSITY, -->

<!--   mode = "WALK", -->

<!--   all_to_all = TRUE # if false, only 1-1 would be calculated -->

<!-- ) -->

<!-- ``` -->

<!-- #### Distances by PT -->

<!-- For Public Transit (`TRANSIT`) you may specify the egress mode, the departure time, and the maximum number of transfers. -->

<!-- ```{r} -->

<!-- #| warning: false -->

<!-- #| code-fold: true -->

<!-- #| eval: false -->

<!-- SURVEYtransit = detailed_itineraries( -->

<!--   r5r_core = r5r_network, -->

<!--   origins = SURVEYsample, -->

<!--   destinations = UNIVERSITY, -->

<!--   mode = "TRANSIT", -->

<!--   mode_egress = "WALK", -->

<!--   max_rides = 1, # The maximum PT rides allowed in the same trip -->

<!--   departure_datetime =  as.POSIXct("20-09-2023 08:00:00", -->

<!--                                  format = "%d-%m-%Y %H:%M:%S"), -->

<!--   all_to_all = TRUE # if false, only 1-1 would be calculated -->

<!-- ) -->

<!-- ``` -->

## Routing distances with `openrouteservice`

We will use the `openrouteservice` r package to estimate the distances using a road network [@openrouteservicer].

::: {.callout-note appearance="simple"}
To properly use the `openrouteservice` package, you need to setup the ORS provider.
See the [setup instructions](software.qmd#ors) for more details.
:::

We will use only respondents with a distance to the university less than 2 km.

```{r}
SURVEYsample = SURVEYgeo |> filter(distance <= 2000)
nrow(SURVEYsample)
```

We need an id (unique identifier) for each survey location, so we can [compare](#compare) them later.

Also, we need to extract the coordinates, from both datasets, to be used in the routing functions of [`openrouteservice::directions()`](https://giscience.github.io/openrouteservice-r/reference/ors_directions.html).

```{r}
# create id columns for both datasets
SURVEYsample = SURVEYsample |>
  mutate(id = c(1:nrow(SURVEYsample))) |>  # from 1 to the number of rows
  mutate(coordinates = st_coordinates(geometry)) # extract coordinates

UNIVERSITY = UNIVERSITY |>
  mutate(id = 1) |> # only one row
  mutate(coordinates = st_coordinates(geometry)) # extract coordinates
```

### Distances by car

Estimate the routes with time and distance by car, from survey locations to University.

::: {.callout-warning appearance="simple"}
This one is not that easy to set-up because the function is prepared to retrieve only one result per request :( So we do a loop function.
:::

```{r orsloop1}
#| eval: false
#| include: true
library(openrouteservice)

SURVEYcar = data.frame() # initial empty data frame

# loop - the origin (i) is the survey location, and the UNIVERSITY is always the same destination
for (i in 1:nrow(SURVEYsample)) {
  ROUTES1 = ors_directions(
    data.frame(
      lon = c(SURVEYsample$coordinates[i, 1], UNIVERSITY$coordinates[1, 1]),
      lat = c(SURVEYsample$coordinates[i, 2], UNIVERSITY$coordinates[1, 2])
    ),
    profile = "driving-car", # or cycling-regular foot-walking
    preference = "fastest", # or shortest
    output = "sf"
  )
  ROUTES1$distance = ROUTES1$summary[[1]]$distance # extract these values from summary
  ROUTES1$duration = ROUTES1$summary[[1]]$duration
  
  SURVEYcar = rbind(SURVEYcar, ROUTES1) # to keep adding in the same df
}

SURVEYcar = SURVEYcar |>
  select(distance, duration, geometry) |> # discard unnecessary variables
  mutate(ID = SURVEYsample$ID) # cbind with syrvey ID
```

### Distances by foot

Repeat the same for `foot-walking`.

```{r orsloop2}
#| code-fold: true
#| eval: false
#| include: true
SURVEYwalk = data.frame() # initial empty data frame

# loop - the origin (i) is the survey location, and the UNIVERSITY is always the same destination
for (i in 1:nrow(SURVEYsample)) {
  ROUTES1 = ors_directions(
    data.frame(
      lon = c(SURVEYsample$coordinates[i, 1], UNIVERSITY$coordinates[1, 1]),
      lat = c(SURVEYsample$coordinates[i, 2], UNIVERSITY$coordinates[1, 2])
    ),
    profile = "foot-walking", # or driving-car cycling-regular cycling-electric
    preference = "fastest", # or shortest
    output = "sf"
  )
  ROUTES1$distance = ROUTES1$summary[[1]]$distance # extract these values from summary
  ROUTES1$duration = ROUTES1$summary[[1]]$duration
  
  SURVEYwalk = rbind(SURVEYwalk, ROUTES1) # to keep adding in the same df
}

SURVEYwalk = SURVEYwalk |>
  select(distance, duration, geometry) |> # discard unnecessary variables
  mutate(ID = SURVEYsample$ID) # cbind with survey ID
```

### Distances by bike

And for `cycling-regular`.

```{r orsloop3}
#| code-fold: true
#| eval: false
#| include: true
SURVEYbike = data.frame() # initial empty data frame

# loop - the origin (i) is the survey location, and the UNIVERSITY is always the same destination
for (i in 1:nrow(SURVEYsample)) {
  ROUTES1 = ors_directions(
    data.frame(
      lon = c(SURVEYsample$coordinates[i, 1], UNIVERSITY$coordinates[1, 1]),
      lat = c(SURVEYsample$coordinates[i, 2], UNIVERSITY$coordinates[1, 2])
    ),
    profile = "cycling-regular", 
    preference = "fastest", # or shortest
    output = "sf"
  )
  ROUTES1$distance = ROUTES1$summary[[1]]$distance # extract these values from summary
  ROUTES1$duration = ROUTES1$summary[[1]]$duration
  
  SURVEYbike = rbind(SURVEYbike, ROUTES1) # to keep adding in the same df
}

SURVEYbike = SURVEYbike |>
  select(distance, duration, geometry) |> # discard unnecessary variables
  mutate(ID = SURVEYsample$ID) # cbind with survey ID
```

```{r importexport1, include=FALSE}
# st_write(SURVEYcar, "../original/routes_car.geojson")
# st_write(SURVEYwalk, "../original/routes_foot.geojson")
# st_write(SURVEYbike, "../original/routes_bike.geojson")

SURVEYcar = st_read("../original/routes_car.geojson", quiet = TRUE)
SURVEYwalk = st_read("../original/routes_foot.geojson", quiet = TRUE)
SURVEYbike = st_read("../original/routes_bike.geojson", quiet = TRUE)
```

```{r}
names(SURVEYcar)
```

::: {.callout-note appearance="simple"}
If we want to know only time and distance, and **not the route** itself, we can use the [`ors_matrix()`](https://giscience.github.io/openrouteservice-r/reference/ors_matrix.html).
:::

## Compare distances {#compare}

We can now compare the euclidean and routing distances that we estimated for the survey locations under 2 km.

```{r}
summary(SURVEYsample$distance) # Euclidean
summary(SURVEYwalk$distance) # Walk
summary(SURVEYcar$distance) # Car
summary(SURVEYbike$distance) # Bike
```

> What can you understand from this results?

```{r}
#| echo: false
distances = data.frame(euclidean = SURVEYsample$distance,
                       walk = SURVEYwalk$distance,
                       car = SURVEYcar$distance,
                       bike = SURVEYbike$distance)
distances = distances |> arrange(euclidean)

# Define the number of observations
n <- nrow(distances)

# Create an empty plot with the appropriate y-axis limits
plot(1:n, distances$euclidean, type = "n", 
     ylim = range(distances), 
     xlab = "Observation (sorted by Euclidean distance)", 
     ylab = "Distance [m]", 
     main = "Distances by Euclidean, Walk, Car, and Bike")

# Add lines for each type of distance
lines(1:n, distances$euclidean, lwd = 2)
lines(1:n, distances$walk, col = "red", lwd = 2)
lines(1:n, distances$car, col = "blue", lwd = 2)
lines(1:n, distances$bike, col = "gold", lwd = 2)

# Optional: Add points for better visibility
points(1:n, distances$euclidean, pch = 16)
points(1:n, distances$walk, col = "red", pch = 16)
points(1:n, distances$car, col = "blue", pch = 16)
points(1:n, distances$bike, col = "gold", pch = 16)

# Add a legend to distinguish the lines
legend("topleft", legend = c("Euclidean", "Walk", "Car", "Bike"),
       col = c("black", "red", "blue", "gold"), 
       lwd = 2, lty = c(1, 2, 3, 4), pch = 16)
```

### Circuity

Compare 1 single route.

```{r}
#| fig-format: png
#| message: false
#| code-fold: true
mapview(SURVEYeuclidean[165,], color = "black") + # 1556 meters
  mapview(SURVEYwalk[78,], color = "red") + # 2126 meters
  mapview(SURVEYcar[78,], color = "blue") + # 2058 meters
  mapview(SURVEYbike[78,], color = "gold") # 2025 meters
```

With this we can see the **circuity** of the routes, a measure of route / transportation efficiency, which is the ratio between the routing distance and the euclidean distance.

The circuity for car (`r round(2058/1556,2)`) is usually lower than for walking (`r round(2126/1556,2)`) or biking, for longer distances, and higher opposite for shorter distances.

## Visualize routes

Visualize with transparency of 30%, to get a clue when they overlay.

```{r}
#| fig-format: png
#| message: false
mapview(SURVEYwalk, alpha = 0.3)
mapview(SURVEYcar, alpha = 0.3, color = "red")
mapview(SURVEYbike, alpha = 0.3, color = "darkgreen")
```

We can also use the [`overline()`](https://docs.ropensci.org/stplanr/reference/overline.html) function from `stplanr` package to break up the routes when they *overline*, and add them up.

```{r}
#| message: false
#| fig-format: png
# we create a value that we can later sum
# it can be the number of trips represented by this route
SURVEYwalk$trips = 1 # in this case is only one respondent per route

SURVEYwalk_overline = stplanr::overline(
  SURVEYwalk,
  attrib = "trips",
  fun = sum
)

mapview(SURVEYwalk_overline, zcol = "trips", lwd = 3)
```

With this we can visually inform on how many people travel along a route, from the survey dataset[^2].

[^2]: Assuming all travel by the shortest path.

::: {.callout-tip appearance="simple"}
## Questions

-   How many people are entering IST by the stairs near *Bar de Civil*?
-   And by the North gate?
-   And from Alameda stairs?
:::
